Early environmental conditions do not impact behavioural flexibility in an invasive and non-invasive lizard species

### 2024-05-30 ms.qmd
Updates

Behavioural flexbility appears to be robust to early exposure to maternal GCs in both _L. delicata_ and _L. guichenoti_. We predicted that prenatal CORT treatment would decrease reversal learning abilities. As such, our results contrast to previous studies that show high GCs levels impaired performance in different types of learning tests [@lemaire_prenatal_2000; @zhu_prenatal_2004;  @farrell_developmental_2016]. Our results are, however, consistent with other experiments where GCs did not impact learning abilities [@szuran_water_1994; @bebus_associative_2016]. One possible reason for why GCs did not affect learning in our study may be because GCs can have hormetic effects on brain structure and learning. For example, in Sprague–Dawley rats, exposure to high doses of CORT, but not low doses, reduced the density of immature neurons and cell proliferation in the dentate gyrus [@DOSEneurobrummelte2010chronic]. It is possible such dose-dependent effects exist and our concentrations failed to elicit changes to the brain that would result in impaired behavioural flexibility. We find this possibility unlikely since, in a previous experiment, we observed that a similar dose affected a multitude of other traits including growth and mortality in _L. delicata_ (unpublished data). Nonetheless, we cannot dismiss this possibility because dose effects can differ between tissues [@lemaire_prenatal_2000]. Conversely, the effects of GCs early in development may vary depending on the cognitive task, as evidenced by studies showing variations in the impact of prenatal GCs on different brain regions [@lemaire_prenatal_2000]. Prenatal GCs can have programmatic effects on HPA axis sensitivity, potentially impacting the lizards’ sensitivity to stressors [@crino_corticosterone_2014]. This altered stress response may influence learning dynamics through hormetic effects of CORT [@du_dynamic_2009], particularly in stressful contexts of which our experiment was not [@taborsky2021towards]. These hypotheses are not mutually exclusive and could be tested in future experiments using multiple doses of prenatal CORT treatment, assessing learning using multiple cognitive tasks and testing the effect of stressors during the learning process.

More striking is the absence of an effect of incubation temperature on learning and behavioural flexibility because most studies have reported better performance in different associative learning and motor tasks when incubated at higher temperatures [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]. Furthermore, those studies that reported cold-incubated lizards to learn faster [@dayananda_incubation_2017; @abayarathna_effects_2020] used incubation temperatures far beyond the natural range in their hot treatments. Our results suggest that behavioural flexibility is robust to temperature in both species of skinks, at least in our study.  **ADD**

Despite sharing life-history traits, only _L. delicata_ has proven successful in colonising various global regions [@baker1979rainbow; @chapple2013divergent; @chapple2015deliinvLHI], and we predicted this may relate to higher behavioural flexibility. Furthermore, we expected behavioural flexibility in this species to be less susceptible to the effect of early environment since this trait can be adaptive in several stages of invasion involving more than one generation. While it will be important to explore replicate invasion events (both successful and unsuccessful) such information is seldom available. Nonetheless, there is precedence in the literature to expect differences between these two species [**REFS**]. Regardless, while we did not find differences in reversal learning between _L. delicata_ and _L. guichenoti_, previous studies have shown that _L. delicata_ more readily explores novel environments than _L. guichenoti_, which can be advantageous during colonisation [@chapple_know_2011]. However, Bezzina _et al._ [@bezzina2014does] found no differences in learning between the two skinks in an associative learning task. Our results align with the latter study and suggest that behavioural flexibility does not differ between these species either. Our results along with others' [@bezzina2014does] suggest that other aspects of their biology are more relevant for _L. delicata_'s success as invaders than learning and behavioural flexibility. Differences in exploratory behaviour is likely to be the key factor explaining why _L. delicata_ and not _L. guichenoti_ is better at colonising new areas [@chapple_know_2011].

In conclusion, our results indicate that lizards performed the reversal learning task, a measure of behavioural flexibility, at similar rates, with no discernible differences between treatments or species. These findings contrast with our initial predictions and suggest that behavioural flexibility in both species is robust to insults (at least for temperature and GCs) during early stages of development. Furthermore, our results indicate that other aspects of the biology of _L. delicata_ may be more influential for their success as invaders than behavioural flexibility alone. Nevertheless, it remains **possible that the effects of the prenatal CORT and temperature on cognition depend on other aspects like the age at which animals are tested or the cognitive task employed; and future studies should explore performance across different cognitive tasks and the effects of both treatments on brain function.**

>>>>

Behavioural flexbility appears to be robust to early exposure to maternal GCs in both _L. delicata_ and _L. guichenoti_. We predicted that prenatal CORT treatment would decrease reversal learning abilities. As such, our results contrast to previous studies that show high GCs levels impaired performance in different types of learning tests [@lemaire_prenatal_2000; @zhu_prenatal_2004;  @farrell_developmental_2016]. Our results are, however, consistent with other experiments where GCs did not impact learning abilities [@szuran_water_1994; @bebus_associative_2016]. One possible reason for why GCs did not affect learning in our study may be because GCs can have hormetic effects on brain structure and learning. For example, in Sprague–Dawley rats, exposure to high doses of CORT, but not low doses, reduced the density of immature neurons and cell proliferation in the dentate gyrus [@DOSEneurobrummelte2010chronic]. It is possible such dose-dependent effects exist and our concentrations failed to elicit changes to the brain that would result in impaired behavioural flexibility. We find this possibility unlikely since, in a previous experiment, we observed that a similar dose affected a multitude of other traits including growth and mortality in _L. delicata_ (unpublished data). Nonetheless, we cannot dismiss this possibility because dose effects can differ between tissues [@lemaire_prenatal_2000]. Conversely, the effects of GCs early in development may vary depending on the cognitive task, as evidenced by studies showing variations in the impact of prenatal GCs on different brain regions [@lemaire_prenatal_2000]. Prenatal GCs can have programmatic effects on HPA axis sensitivity, potentially impacting the lizards’ sensitivity to stressors [@crino_corticosterone_2014]. This altered stress response may influence learning dynamics through hormetic effects of CORT [@du_dynamic_2009], particularly in stressful contexts of which our experiment was not [@taborsky2021towards]. These hypotheses are not mutually exclusive and could be tested in future experiments using multiple doses of prenatal CORT treatment, assessing learning using multiple cognitive tasks, and testing the effect of stressors during the learning process.

More striking is the absence of an effect of incubation temperature on learning and behavioural flexibility because most studies have reported better performance in different associative learning and motor tasks when incubated at higher temperatures [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]. Furthermore, those studies that reported cold-incubated lizards to learn faster [@dayananda_incubation_2017; @abayarathna_effects_2020] used incubation temperatures far beyond the natural range in their hot treatments. Our results suggest that behavioural flexibility is robust to temperature in both species of skinks, at least in our study. It could be arguable whether the differences in our incubation temperatures were extreme enough to elicit significant differences in learning performance. However, the mean difference between the temperatures used in our experiment  were similar to those in previous studies [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]. In lizards, disparities in learning caused by the early thermal environment are associated to changes in neural density [@amiel_effects_2017], however, several studies have shown that the effects of temperature on brain structure and physiology are region dependent [@coomber_independent_1997; @sakata_neural_2000; @amiel_effects_2017]. It is possible that the temperatures used in our experiment did not affect the brain regions involved in coding reversal learning. Future studies should explore the effects of thermal environment on brain structure and function in these species to understand the mechanisms underlying the effects of temperature on learning. Furthermore, testing learning across different cognitive tasks could provide a more comprehensive understanding of the influence of temperature on cognition. In our analyses, sex was not included as a factor because the lizads employed were not considered sexually mature, and because previous studies did not find any differences on learning [@szabo2019sex], or on the effects of temperature on learning [@amiel_hotter_2012; @clark_colour_2014]. However, some studies evidence interactive effects of incubation temperature and sex on brain physiology in other species of reptiles [@coomber_independent_1997; @sakata_neural_2000]. In addition, the species employed by @amiel_hotter_2012, @amiel_egg_2014, and @clark_colour_2014, is known to sex-reverse - some individuals can have female chromosomes but male phenotypes - under cold temperatures [@dissanayake2021effects]; which was not accounted for in these experiments [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]. As such, a potential temperature-sex interaction could be impacting our results, and future studies should explore the interactive effects of sex and incubation temperature on cognitive performance.

Despite sharing life-history traits, only _L. delicata_ has proven successful in colonising various global regions [@baker1979rainbow; @chapple2013divergent; @chapple2015deliinvLHI], and we predicted this may relate to higher behavioural flexibility. Furthermore, we expected behavioural flexibility in this species to be less susceptible to the effect of early environment since this trait can be adaptive in several stages of invasion involving more than one generation. While it will be important to explore replicate invasion events (both successful and unsuccessful) such information is seldom available. Nonetheless, there is precedence in the literature to expect differences between these two species [@chapple_know_2011; @chapple2013biosecurity]. Regardless, while we did not find differences in reversal learning between _L. delicata_ and _L. guichenoti_, previous studies have shown that _L. delicata_ more readily explores novel environments than _L. guichenoti_, which can be advantageous during colonisation [@chapple_know_2011]. However, @bezzina2014does found no differences in learning between the two skinks in an associative learning task. Our results align with the latter study and suggest that behavioural flexibility does not differ between these species either. Our results along with others' [@bezzina2014does] suggest that other aspects of their biology are more relevant for _L. delicata_'s success as invaders than learning and behavioural flexibility. Differences in exploratory behaviour is likely to be the key factor explaining why _L. delicata_ and not _L. guichenoti_ is better at colonising new areas [@chapple_know_2011].

In conclusion, our results indicate that lizards performed the reversal learning task, a measure of behavioural flexibility, at similar rates, with no discernible differences between treatments or species. These findings contrast with our initial predictions and suggest that behavioural flexibility in both species is robust to insults (at least for temperature and GCs) during early stages of development. Furthermore, our results indicate that other aspects of the biology of _L. delicata_ may be more influential for their success as invaders than behavioural flexibility alone. Nevertheless, future studies should investigate the potential influence of prenatal CORT and temperature on cognition, considering other factors such as the type of cognitive task. It is also imperative to explore the effects of these treatments on brain function comprehensively. Additionally, further research is essential to determine which aspects of _L. delicata_’s biology contribute more significantly to their success as invaders than behavioral flexibility alone. Investigating these areas will provide a better understanding of the cognitive and biological mechanisms underlying the adaptability and invasive potential of both species.
